Type theory with single substitutions

### 2025-02-21 cpp/p.tex
links

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

@@

\subparagraph*{B-systems.} B-systems introduced by Voevodsky \cite{voevodsky2014bsystems}
are an intrinsic, essentially
algebraic presentation of type theory using single
substitutions. B-systems relate to our SSC as essentially
algebraic theories relate to generalised algebraic theories, or
sets with a map into $I$ relate to indexed families over $I$
\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
that we have fewer and less general equations, resulting in the fact
that we have more models than B-systems. However in the syntax of our
theory, all the rules of B-systems are admissible. We describe the
relationship in more detail in Appendix \ref{app:bsys}.
%
Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space.
%Their equivalence result holds even without the presence of function types.
%In the presence of function space,
The same equivalence holds for the simply typed version of our
substitution calculus (TODO link).

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda. The formalisation is available at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
with postulated coercion rule (transport rule) and postulated QIITs
with computation rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
overloading extensively, for example instantiation $\blank[\blank]$ is
overloaded for types and terms. For isomorphisms, we use the notation
$f:X\cong Y:g$ meaning $f:X\ra Y$ and $g :Y\ra X$ together with a
$\beta$ equality $f\,(g\,y) = y$ for all $y$, and an $\eta$ equality
saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
and $g$ as a constructor).

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the definition of our basic type theory. For
reference, we list the same rules again in Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules.\footnote{In \cite{DBLP:conf/fscd/KaposiX24},
the same structural rules suffice to describe the first-order theory for any SOGAT.}
For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \rlap{$
  \top : \Ty\,\Gamma\,0 \hspace{3em}
  \top[] : \top[\gamma] = \top\hspace{3em}
  \top\eta : t = \tt \hspace{3em}
  \tt : \Tm\,\Gamma\,\top \hspace{3em}
  \tt[] : \tt[\gamma] = \tt
      $} \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into an isomorphism, and did not list
  substitution law for $\fst$, $\snd$ as they are derivable (see
  below).
\end{definition}

@@

\begin{definition}[CwF]\label{def:cwf}
  A category with $\N$-many families is defined as a category (objects
  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
  action on morphisms $\blank[\blank]$); for each $i$ a locally
  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
  with an isomorphism
  $
  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
  $
  natural in $\Delta$).
\end{definition}
In the rest of this section, we work with the SSC syntax denoted
$\Con$, $\Sub$, $\Ty$, $\Tm$, and so on, towards the goal of defining
all CwF operations and equations.

@@

To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we
define telescopes as an inductive type together with a recursive
operation to append telescopes to contexts.
\begin{alignat*}{10}
  & \Tel && : \Con\ra\Set   && \rlap{$\blank+\blank : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con$} \\
  & \diamond && : \Tel\,\Gamma && \Gamma+\diamond && := \Gamma \\
  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma \hspace{3em} && \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
\end{alignat*}
We then define a lifting operation over any telescope mutually with an
instantiation operation on telescopes by recursion on telescopes.
\[
\blank[\blank] : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \hspace{3em}
\blank^{+^\Omega} : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
\]

@@

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$ as follows.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers such as
$\lam$. Notably, we do not state that instantiated types/terms are
$\alpha$-normal (except variables). It is important to truncate the
$\alpha$-normal predicate to be a proposition to ensure that it preserves the
$\beta$/$\eta$ conversion rules.

We prove that the $\alpha$-normal predicate holds for $\alpha$-normal types and
terms instantiated with $\alpha$-normal substitutions. However, this needs to be
proved separately for weakenings and $\alpha$-normal single substitutions for
the induction to be structural. We define predicates for weakening and
$\alpha$-normal single substitutions as follows.
\begin{itemize}
  \item $\p$ is a weakening.
  \item $\gamma^+$ is a weakening if $\gamma$ is a weakening.
  \item $\langle a \rangle$ is an $\alpha$-normal single substitution if $a$ is an
  $\alpha$-normal term.
  \item $\gamma^+$ is an $\alpha$-normal single substitution if $\gamma$ is an
  $\alpha$-normal single substitution.
\end{itemize}
We define $\alpha$-normal substitutions to be the disjoint union of weakenings
and $\alpha$-normal single substitutions.

\begin{lemma}
  \label{thm:alpha}
  The $\alpha$-normal predicate holds for any type and term.
\end{lemma}
\begin{proof}
  By induction on the syntax, $\alpha$-normalizing the substitutions
  at the same time. Note that induction on the syntax refers to the
  elimination principle of the corresponding QIIT.
\end{proof}

Instead of doing induction on $\alpha$-normal forms for each of equations 
(\ref{eq:lifted1})--(\ref{eq:lifted4}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if for any $A:\Ty\,\Gamma$, $A[\gamma_0] = A[\gamma_1]$, and
  for any variable $x:\Tm\,\Gamma\,A$, $x[\gamma_0] = x[\gamma_1]$,
  then:
  \begin{itemize}
    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend on the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
difficult to define the instantiation operation for $\Tms$ directly, so we also define
a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
convert it into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con && \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{1.9em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated below. On the
right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
then replace the variables of $B$ one by one using the $a_i$'s, which are
well-typed because of the previous weakening.
\[
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = 
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\]
All CwF operations and equations can be defined with $\Tms$ by induction, using
Lemma \ref{thm:lifting} to avoid further induction on the syntax.

\begin{problem}
  Contexts, types, and terms in the SSC syntax are isomorphic to the
  corresponding sorts in the CwF syntax. In addition $\Tms$ is isomorphic to CwF
  substitutions.
\end{problem}
\begin{proof}[Construction]\leavevmode
  \begin{description}
    \item[$\Rightarrow$] By recursion on the syntax, SSC operations can be
    trivially interpreted by CwF operations.
    \item[$\Leftarrow$] By recursion on the syntax, using $\Tms$ to interpret
    parallel substitutions.
  \end{description}
  The round-trips are proven by induction on the syntax.
\end{proof}
This also implies that the initial SSC is the initial CwF.

@@

% TODO: named CwFs would be even closer to usual syntax.

>>>>

%\supplement{Formalisation: \url{https://szumixie.github.io/single-subst}} % TODO: add this back

@@

\subparagraph*{B-systems.} B-systems introduced by Voevodsky \cite{voevodsky2014bsystems}
are an intrinsic, essentially
algebraic presentation of type theory using single
substitutions. B-systems relate to our SSC as essentially
algebraic theories relate to generalised algebraic theories, or
sets with a map into $I$ relate to indexed families over $I$
\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
that we have fewer and less general equations, resulting in the fact
that we have more models than B-systems. However in the syntax of our
theory, all the rules of B-systems are admissible. We describe the
relationship in more detail in Appendix \ref{app:bsys}.
%
Kaposi and Luksa \cite{luksa} defined a telescopic SSC for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs.
%Their equivalence result holds even without the presence of function types.
%In the presence of function space,
If we have function space, the same equivalence holds for the simply typed version of our
substitution calculus (\href{https://szumixie.github.io/single-subst/STT.Contextual.html}{\Agda}).

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Definition \ref{def:tt} and Section
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda. The formalisation is available online
(\href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}). The Agda
logos next to definitions/theorems are links to their formalised counterparts. We use a strict Prop-valued
\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
coercion rule (transport rule) and postulated QIITs with computation
rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
overloading extensively, for example instantiation $\blank[\blank]$ is
overloaded for types and terms. For isomorphisms, we use the notation
$f:X\cong Y:g$ meaning $f:X\ra Y$ and $g :Y\ra X$ together with a
$\beta$ equality $f\,(g\,y) = y$ for all $y$, and an $\eta$ equality
saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
and $g$ as a constructor).

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Syntax.html}{\Agda})]\label{def:tt}
This concludes the definition of our basic type theory. For
reference, we list the same rules again in Appendix \ref{app:tt}.
\end{definition}

As a sanity check for our notion of SSC-model, we defined the standard (metacircular\slash set/type) model \cite{DBLP:conf/popl/AltenkirchK16} of the
SSC-based calculus in Agda (\href{https://szumixie.github.io/single-subst/TT.SSC.Standard.html}{\Agda}).

The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules.\footnote{In \cite{DBLP:conf/fscd/KaposiX24},
the same structural rules suffice to describe the first-order theory for any SOGAT.}
For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \rlap{$
  \top : \Ty\,\Gamma\,0 \hspace{3em}
  \top[] : \top[\gamma] = \top\hspace{3em}
  \top\eta : t = \tt \hspace{3em}
  \tt : \Tm\,\Gamma\,\top \hspace{3em}
  \tt[] : \tt[\gamma] = \tt
      $} \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into an isomorphism, and did not list
  substitution law for $\fst$, $\snd$ as they are derivable (see
  below).
\end{definition}

@@

\begin{definition}[CwF (see \href{https://szumixie.github.io/single-subst/TT.CwF.Syntax.html}{\Agda})]\label{def:cwf}
  A category with $\N$-many families is defined as a category (objects
   $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
  action on morphisms $\blank[\blank]$); for each $i$ a locally
  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
  with an isomorphism
  $
  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
  $
  natural in $\Delta$).
\end{definition}
In the rest of this section, we work with the SSC syntax denoted
$\Con$, $\Sub$, $\Ty$, $\Tm$, and so on, towards the goal of defining
all CwF operations and equations.

@@

To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we
define telescopes as an inductive type together with a recursive
operation to append telescopes to contexts (\href{https://szumixie.github.io/single-subst/TT.SSC.Tel.html#Tel}{\Agda}).
\begin{alignat*}{10}
  & \Tel && : \Con\ra\Set   && \rlap{$\blank+\blank : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con$} \\
  & \diamond && : \Tel\,\Gamma && \Gamma+\diamond && := \Gamma \\
  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma \hspace{3em} && \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
\end{alignat*}
We then define a lifting operation over any telescope mutually with an
instantiation operation on telescopes by recursion on telescopes (\href{https://szumixie.github.io/single-subst/TT.SSC.Tel.html#_[_]\%E1\%B5\%80\%CB\%A1}{\Agda}).
\[
\blank[\blank] : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \hspace{3em}
\blank^{+^\Omega} : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
\]

@@

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$ as follows (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#Var}{\Agda}).
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers such as
$\lam$. Notably, we do not state that instantiated types/terms are
$\alpha$-normal (except variables). It is important to truncate the
$\alpha$-normal predicate to be a proposition to ensure that it preserves the
$\beta$/$\eta$ conversion rules.

We prove that the $\alpha$-normal predicate holds for $\alpha$-normal types and
terms instantiated with $\alpha$-normal substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html\#[]\%E1\%B4\%BA\%E1\%B4\%BE}{\Agda}). However, this needs to be
proved separately for weakenings and $\alpha$-normal single substitutions for
the induction to be structural. We define predicates for weakening (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#Wk}{\Agda}) and
$\alpha$-normal single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#NSSub}{\Agda}) as follows.
\begin{itemize}
  \item $\p$ is a weakening.
  \item $\gamma^+$ is a weakening if $\gamma$ is a weakening.
  \item $\langle a \rangle$ is an $\alpha$-normal single substitution if $a$ is an
  $\alpha$-normal term.
  \item $\gamma^+$ is an $\alpha$-normal single substitution if $\gamma$ is an
  $\alpha$-normal single substitution.
\end{itemize}
We define $\alpha$-normal substitutions to be the disjoint union of weakenings
and $\alpha$-normal single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#NSub}{\Agda}).

\begin{lemma}[\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html\#norm\%E1\%B5\%80}{\Agda}]
  \label{thm:alpha}
  The $\alpha$-normal predicate holds for any type and term.
\end{lemma}
\begin{proof}
  By induction on the syntax, $\alpha$-normalizing the substitutions
  at the same time. Note that induction on the syntax refers to the
  elimination principle of the corresponding QIIT.
\end{proof}

Instead of doing induction on $\alpha$-normal forms for each of equations 
(\ref{eq:lifted1})--(\ref{eq:lifted4}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html#Sub*}{\Agda}), we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html}{\Agda}).
\begin{lemma}
  \label{thm:lifting}
  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if for any $A:\Ty\,\Gamma$, $A[\gamma_0] = A[\gamma_1]$, and
  for any variable $x:\Tm\,\Gamma\,A$, $x[\gamma_0] = x[\gamma_1]$,
  then:
  \begin{itemize}
    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5574}{\Agda})} $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5713}{\Agda})} $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5874}{\Agda})} $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend on the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
difficult to define the instantiation operation for $\Tms$ directly, so we also define
a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
convert it into a sequence of instantiations of single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html\#Tms}{\Agda}).
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con && \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{1.9em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html\#559}{\Agda}) is illustrated below. On the
right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
then replace the variables of $B$ one by one using the $a_i$'s, which are
well-typed because of the previous weakening.
\[
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = 
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\]
All CwF operations and equations can be defined with $\Tms$ by induction, using
Lemma \ref{thm:lifting} to avoid further induction on the syntax (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html}{\Agda}).

\begin{problem}[\href{https://szumixie.github.io/single-subst/TT.Isomorphism.html}{\Agda}]
  Contexts, types, and terms in the SSC syntax are isomorphic to the
  corresponding sorts in the CwF syntax. In addition $\Tms$ is isomorphic to CwF
  substitutions.
\end{problem}
\begin{proof}[Construction]\leavevmode
  \begin{description}
    \item[$\Rightarrow$] By recursion on the syntax, SSC operations can be
    trivially interpreted by CwF operations.
    \item[$\Leftarrow$] By recursion on the syntax, using $\Tms$ to interpret
    parallel substitutions.
  \end{description}
  The round-trips are proven by induction on the syntax.
\end{proof}
This also implies that the initial SSC is the initial CwF.

@@

% named CwFs would be even closer to usual syntax.
